---
title: "Reproducible Research: Peer Graded Assignment"
author: "Sheila Amoafo"
date: "09/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading and processing the data.

First off, the necessary packages are loaded.

```{r,echo=TRUE}
library(ggplot2)
library(dplyr)
```

Then the data is loaded and examined.

```{r,echo=TRUE}
activity<-read.csv("activity.csv")
summary(activity)
head(activity)
```

Process the data into a suitable format

```{r,echo=TRUE}
#remove NAs
actNA<-is.na(activity$steps)
actnew<-activity[!actNA,]
head(actnew)
```







# Mean total number of steps taken per day

## 1. Total number of steps taken per day

```{r}
#Select the steps and date column and group by date
bydate<-actnew %>% select(date,steps) %>% group_by(date) %>% summarise(steps=sum(steps))
head(bydate)
```


## 2. Histogram of total number of steps taken each day

```{r}
#Plot the histogram
hist(bydate$steps, xlab = "Daily Steps", main = "Total Steps per Day",col = "pink",breaks = 20)
```


## 3. Mean and Median of total number of steps taken per day

```{r}
#calculating the mean of the total number of steps
mean(bydate$steps)

#calculating the median of the total number of steps
median(bydate$steps)
```







# Average Daily Activity Pattern

## 4. Time series plot of the 5-minute interval and the average number of steps

```{r}
#Select the interval and average steps columns
byinterval<- actnew %>% select(interval,steps) %>% group_by(interval) %>% summarise(steps = mean(steps))

#plot the time series plot
ggplot(byinterval, aes(x=interval,y=steps))+geom_line(color= "magenta")+labs(title = "Time Series Plot", x = "Interval", y = "Average Steps")
```


## 5. 5-minute interval containing the maximum number of steps on average

```{r}
byinterval[which(byinterval$steps==max(byinterval$steps)),]
```







# Imputing missing values

## 6. Total number of missing values in the dataset

```{r}
#the total number of NAs was stored in the object actNA under loading and processing the data
sum(actNA)
```


## 7. Filling in missing Values

```{r}
#My strategy is to fill the NAs with the mean of the interval. Hence i wrote a function that performs that function;
fillNA<- function(x){replace(x,is.na(x),mean(x,na.rm = TRUE))}
```


## 8. Create new dataset equal to original but without NAs

```{r}
#Apply the function and head the result to view the first 6 rows of the  new dataset
activitynew<-activity %>% group_by(interval) %>% mutate(steps= fillNA(steps))
head(activitynew)
```


## 9. Histogram of total number of steps taken each day of the new dataset

```{r}
#Select the date and steps columns from the new data
maindata<-activitynew %>% select(date, steps) %>% group_by(date) %>% summarise(steps= sum(steps))

#view part of the data
head(maindata,20)

#plot the histogram
hist(maindata$steps,xlab = "Steps", main = "Total Daily Steps", col = "pink", breaks = 20)
```


## 10. Mean and Median of total number of steps taken per day of the new dataset

```{r}
#mean of total number of steps
mean(maindata$steps)

#median of total number of steps
median(maindata$steps)
```


## 11. Compare and determine impact of adding the NAs to the mean and median to the old and new dataset

```{r}
#old mean
mean(bydate$steps)

#new mean
mean(maindata$steps)
```

There was no noticable difference in the means

```{r}
#old median
median(bydate$steps)

#new median
median(maindata$steps)
```

There was a slight increase in the median







# Are there differences in activity patterns between weekdays and weekends

## 12. Factor variable indicating weekdays and weekends

```{r}
maindata$date <- as.Date(maindata$date)
#extracting the weekdays
maindata$weekday<-weekdays(maindata$date)

#Extracting the weekends
maindata$weekend<- ifelse(maindata$weekday=="Saturday" | maindata$weekday=="Sunday", "weekend", "weekday")
```


## 13. Panel plot containing time series plot of new data

```{r}
weekendweekday<-aggregate(maindata$steps, by=list(maindata$weekend,maindata$weekday),mean)
names(weekendweekday)<-c("steps","weekday","interval")

ggplot(weekendweekday,aes(x = interval,y = steps,color= weekday))+geom_line()+labs(title = "Comparison of Average Number of Steps")